Financialisation of Commodity Markets

### 2024-12-04 docs/paper.qmd
paper edits
refactor make_table function + introduce cross correlation analysis in methods section.

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Co-movement
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear
  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

```{r}
#| label: globals
#| include: false

@@

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

urls <- c(
  "https://www.dropbox.com/scl/fi/n7t8g3hn32d8e5gf2rcti/references.bib?rlkey=xy251n44q3y0t3omxpqk0dvyf&dl=1",
  "https://www.dropbox.com/scl/fi/c46nb5xpxwgvjdspla9p5/style.css?rlkey=d666hnaxod2stvkxel33or0cg&dl=1"
)
file_names <- c("bibliography.bib", "style.css")
purrr::walk2(urls, file_names, function(url, file_name) download.file(url, file_name))

@@

make_table <- function(table_name, columns_to_collapse_rows_for, colnames, align, bold_columns, coloured_columns){
  get_results(tables, table_name) %>% dplyr::mutate(dplyr::across(dplyr::where(is.factor), as.character)) %>%
  collapse_rows(columns_to_collapse_rows_for) %>%
  kbl(booktabs = T, longtable = T, col.names = colnames, align = align) %>%
  kable_styling(latex_options = "striped", font_size = 7L) %>%
  row_spec(0L, bold = TRUE, color = "black") %>%
  column_spec(bold_columns, bold = TRUE) %>% column_spec(coloured_columns, color = "#4285f4")
}
```

@@

For each period, the weekly return series for the individual US-traded assets are regressed on both the relative change on their respective CHP series and that on the cross-sectional CHP independently. The resulting $R^{2}$ values are then averaged by sector and sub-sector. Additionally, pairwise daily return correlations are computed for each combination of assets across the periods of interest and for each individual year within the full sample period. These correlation coefficients are then averaged across the entire cross-section of assets, as well as by country, sector, and sub-sector. For each period, the coefficients are also calculated and averaged across phases of low and high cross-sectional CHP.

This approach sheds light on the evolving interconnectedness among commodity assets as the complex transitions from traditional market dynamics---where commodity producers and speculators had distinct roles---to a more intricate landscape shaped by the influx of financial investors.

@@

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more cohesive collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-inner-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-combined.   
We break down these average correlations by years in @tbl-correlations-inner-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

@@

colnames <- c("asset", "regime", "estimate", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))

make_table("stats - combined", c("asset", "regime"), colnames, align, 1L:7L, 4L:7L)
```

@@

colnames <- c("asset", "period", "moment", "dominant regime", "p-value", "significance")
align <- make_series_of_repeated_chars(list(l = 4L, r = 2L))

make_table("regime difference tests", c("asset", "period"), colnames, align, 1L:4L, 5L:6L)
```

@@

colnames <- c("regressor", "sector", "sub-sector", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))

make_table("regressions - US returns ~ US CHP", c("regressor", "sector"), colnames, align, 1L:7L, 4L:7L)
```

@@

colnames <- c(sectors, "regime", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 5L))

make_table("correlations - inner - periods", c("country", "sector", "subsector"), colnames, align, 1L:8L, 5L:8L)
```

@@

colnames <- c("portfolios", "regime", periods)
align <- make_series_of_repeated_chars(list(l = 1L, r = 5L))

make_table("correlations - cross - periods", c("portfolios", "regime"), colnames, align, 1L:2L, 3L:6L)
```

@@

colnames <- c(sectors, "decade", 0L:9L)
align <- make_series_of_repeated_chars(list(l = 4L, r = 10L))

make_table("correlations - inner - years", c("country", "sector", "subsector"), colnames, align, 1L:14L, 5L:14L) %>%
  landscape()
```

@@

colnames <- c(sectors, periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))

make_table(
  "regressions - all returns ~ market index", c("country", "sector", "subsector"), 
  colnames, align, 1L:7L, 4L:7L
  )
```

@@

colnames <- c(sectors, "factor", "regime", periods)
align <- make_series_of_repeated_chars(list(l = 5L, r = 4L))

make_table(
  "regressions - all returns ~ factors", c("country", "sector", "subsector", "factor"), 
  colnames, align, 1L:9L, 6L:9L
  ) %>% landscape()
```

@@

colnames <- c("asset", "country", "sector", "sub-sector")
align <- make_series_of_repeated_chars(list(l = 4L))

make_table("assets taxonomy", c("country", "sector", "subsector"), colnames, align, 1L, 2L:4L)
```

>>>>

---
title: Financialisation of Commodity Markets
subtitle: Co-movement Behind-the-Scenes
author:
  - name: Devraj Basu
    email: devraj.basu@strath.ac.uk
    affiliations: 
        - id: strathclyde
          name: University of Strathclyde Business School
          department: Accounting & Finance
          address: 199 Cathedral Street
          city: Glasgow
          state: Scotland
          postal-code: G4 0QU
    attributes:
        corresponding: true
  - name: Olivier Bauthéac
    email: olivier.bautheac@strath.ac.uk
    affiliations:
        - ref: strathclyde
abstract: |
  In the early 2000s institutional investors entered the commodity futures markets en masse with passive, long only, index type positions in sharp contrast with those typically assumed by traditional expert participants. A heated public debate soon erupted over the perceived consequences of the phenomenon---commonly referred to as "financialisation"---and, in response to immediate policy concerns, the matter was thrust onto the academic sphere as a burning issue. With the benefit of hindsight it now seems that the academic debate was framed rather narrowly, used contentious research methods, and eventually led to regulatory changes that were therefore perhaps unwarranted. In contrast, we take a broader approach where we consider a large cross-section of liquid commodities as suggested by the nature of financialisation that comprehends commodity futures as an asset class. We examine the cross-sectional interconnectedness of these assets and uses a bespoke asset pricing factors based framework to study the issue through the lens of co-movement. We find that the phenomenon had ontological consequences for the commodity complex with its impact extending beyond the mechanical effects induced by indexation. The onset of the financial crisis and the monetary policy regimes that followed, on the other hand, seem to have set off a motion of reversion to legacy pre-financialisation fundamentals.
keywords: 
  - Commodity markets
  - Financialisation
  - Co-movement
  - Asset pricing
date: last-modified
bibliography: bibliography.bib
code_repository: https://github.com/bautheac/financialisation-co-movement
format:
  elsevier-pdf:
    keep-tex: true
    journal:  
      name: Journal of Futures Markets
      formatting: preprint
      model: 3p
      cite-style: authoryear

  html:
    grid:
      body-width: 1000px
    toc: true
    toc_depth: 3
    toc-title: Contents
    css: style.css
    code-links:
      - text: github repository
        icon: github
        href: https://github.com/bautheac/financialisation-co-movement
---

```{r}
#| label: fetch
#| include: false
#| cache: true

urls <- c(
  "https://www.dropbox.com/scl/fi/n7t8g3hn32d8e5gf2rcti/references.bib?rlkey=xy251n44q3y0t3omxpqk0dvyf&dl=1",
  "https://www.dropbox.com/scl/fi/c46nb5xpxwgvjdspla9p5/style.css?rlkey=d666hnaxod2stvkxel33or0cg&dl=1"
)
file_names <- c("bibliography.bib", "style.css")
purrr::walk2(urls, file_names, function(url, file_name) download.file(url, file_name))
```

```{r}
#| label: globals
#| include: false
#| cache: false

knitr::opts_chunk$set(cache = FALSE, echo = FALSE, message = FALSE, warning = FALSE)

@@

@@

make_table <- function(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns){
  get_results(tables, table_name) %>% dplyr::mutate(dplyr::across(dplyr::where(is.factor), as.character)) %>%
    collapse_rows(columns_to_collapse_rows_for) %>%
    kbl(booktabs = T, longtable = longtable, col.names = colnames, align = align) %>%
    kable_styling(latex_options = "basic", font_size = 7L) %>%
    row_spec(0L, bold = TRUE, color = "black") %>%
    column_spec(bold_columns, bold = TRUE) %>% column_spec(coloured_columns, color = "#4285f4")
}
```

@@

For each period, the weekly return series for the individual US-traded assets are regressed on both the relative change on their respective CHP series and that on the cross-sectional CHP independently. The resulting $R^{2}$ values are then averaged by sector and sub-sector. Additionally, pairwise daily return correlations are computed for each combination of assets across the periods of interest and for each individual year within the full sample period. These correlation coefficients are then averaged across the entire cross-section of assets, as well as by country, sector, and sub-sector. For each period, the coefficients are also calculated and averaged independently over phases of low and high cross-sectional CHP. We further form equally weighted portfolios by country, by grouping together the US and GB traded assets independently, as well a
by sector and sub-sector for the US traded assets. For each period we compute the correlation coefficient between the two country portfolios as well as the average pairwise correlation coefficients between the sector and sub-sector portfolios; these are computed over the entire period as well as over phases of low and high cross-sectional CHP.  
This approach sheds light on the evolving interconnectedness among commodity assets as the complex transitions from traditional market dynamics---where commodity producers and speculators had distinct roles---to a more intricate landscape shaped by the influx of financial investors.

@@

Financialisation was a global phenomenon, through the channel of index investment it simultaneously influenced the entire spectrum of liquid commodities, eventually spurring the transformation of commodity markets from a collection of idiosyncratic markets into a more cohesive collection akin to an asset class. In this context, interconnectedness emerges as a crucial aspect that warrants thorough examination.   
We strive to address this issue by first examining the average returns pairwise correlations between commodities across the entire cross-section of the commodity assets considered in the study as well as by country, sector and sub-sector. @tbl-correlations-inner-periods shows the results for the four periods of interest with the results reported for the entire period as well as for periods of **backwardation** and **contango**. For the entire sample the average pairwise correlation jumps from 0.07 in the first period to 0.16 over financialisation. This finding indicates higher interdependence within the full cross-section of commodities starting with the financialisation period, consistent with the empirical findings in @fryMcKibbin_evolution_2023 and @mayer_financialization_2017 as well as with the theoretical model of @basak_model_2016. The sector analysis sheds light on the driving force behind these changes. The US metals see the sharpest increase in pairwise correlations going from 0.22 over the past period to 0.56 over financialisation while the three other sectors see a much smaller percentage increase during financialisation relative to the past period. These results thus suggest that the US metals sector was the driving force behind increased average pairwise correlations during financialisation, a new observation to the best of our knowledge. The pattern of correlations is very similar across phases of **backwardation** and **contango** for all the sectors and sub-sectors over the past and financialisation periods which indicates that increased commodity interdependence over financialisation was not a phase driven phenomenon in the same way that individual commodity returns are, as demonstrated in @tbl-stats-combined.   
We break down these average correlations by year in @tbl-correlations-inner-years. For the full sample of assets, the average pairwise correlations remain in the 0.06–0.08 range from 1998 through 2003, before suddenly increasing to 0.12 from 2003 to 2004. This suggests 2004 as a distinct break point for the onset of financialisation, as noted in several studies [@baker_financialization_2021; @tang_index_2012]. We further observe that the US metals appear to have played a major role in this as the increase seems driven by the sector which average pairwise correlation more than doubles (0.49 vs 0.22) over the period.

@@

\newgeometry{margin=.75in}

@@

table_name <- "stats - combined"
columns_to_collapse_rows_for <- c("asset", "regime")
colnames <- c("asset", "regime", "estimate", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))
longtable <- TRUE
bold_columns <- 1L:3L
coloured_columns <- 4L:7L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```

@@

table_name <- "regime difference tests"
columns_to_collapse_rows_for <- c("asset", "period")
colnames <- c("asset", "period", "moment", "dominant regime", "p-value", "significance")
align <- make_series_of_repeated_chars(list(l = 4L, r = 2L))
longtable <- TRUE
bold_columns <- 1L:4L 
coloured_columns <- 5L:6L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```

@@

table_name <- "regressions - US returns ~ US CHP"
columns_to_collapse_rows_for <- c("regressor", "sector")
colnames <- c("regressor", "sector", "sub-sector", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))
longtable <- TRUE
bold_columns <- 1L:7L
coloured_columns <- 4L:7L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```

@@

table_name <- "correlations - inner - periods"
columns_to_collapse_rows_for <- c("country", "sector", "subsector")
colnames <- c(sectors, "regime", periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 5L))
longtable <- TRUE
bold_columns <- 1L:8L 
coloured_columns <- 5L:8L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```

table_name <- "correlations - cross - periods"
columns_to_collapse_rows_for <- c("portfolios", "regime")
colnames <- c("portfolios", "regime", periods)
align <- make_series_of_repeated_chars(list(l = 1L, r = 5L))
longtable <- TRUE
bold_columns <- 1L:2L
coloured_columns <- 3L:6L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```

@@

```{r}
#| eval: false
#| label: tbl-experimental
#| tbl-cap: "lorem ipsum dolor sit amet consectetur adipiscing elit sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident sunt in culpa qui officia deserunt mollit anim id est laborum. Lorem ipsum dolor sit amet consectetur adipiscing elit sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident sunt in culpa qui officia deserunt mollit anim id est laborum."
#| tbl-subcap: 
#|     - "Average pairwise inner correlations"
#|     - "Average pairwise cross correlations"
#| layout-nrows: 2

table1_format <- list(
  table_name = "correlations - inner - periods",
  columns_to_collapse_rows_for = c("country", "sector", "subsector"),
  longtable = FALSE,
  colnames = c(sectors, "regime", periods),
  align = make_series_of_repeated_chars(list(l = 3L, r = 5L)),
  bold_columns = 1L:8L, 
  coloured_columns = 5L:8L
  )
table2_format <- list(
  table_name = "correlations - cross - periods",
  columns_to_collapse_rows_for = c("portfolios", "regime"),
  longtable = TRUE,
  colnames = c("portfolios", "regime", periods),
  align = make_series_of_repeated_chars(list(l = 1L, r = 5L)),
  bold_columns = 1L:2L, 
  coloured_columns = 3L:6L
)
formats <- list(table1_format, table2_format)

get_results(tables, table1_format$table_name) %>% dplyr::mutate(dplyr::across(dplyr::where(is.factor), as.character)) %>%
    collapse_rows(table1_format$columns_to_collapse_rows_for) %>%
    kable(booktabs = T, longtable = table1_format$longtable, col.names = table1_format$colnames, align = table1_format$align, escape = FALSE) %>%
    kable_styling(font_size = 7L) %>%
    row_spec(0L, bold = TRUE, color = "black") %>%
    column_spec(table1_format$bold_columns, bold = TRUE) %>% column_spec(table1_format$coloured_columns, color = "#4285f4")

get_results(tables, table2_format$table_name) %>% dplyr::mutate(dplyr::across(dplyr::where(is.factor), as.character)) %>%
    collapse_rows(table2_format$columns_to_collapse_rows_for) %>%
    kable(booktabs = T, longtable = table2_format$longtable, col.names = table2_format$colnames, align = table2_format$align, escape = FALSE) %>%
    kable_styling(font_size = 7L) %>%
    row_spec(0L, bold = TRUE, color = "black") %>%
    column_spec(table2_format$bold_columns, bold = TRUE) %>% column_spec(table2_format$coloured_columns, color = "#4285f4")
```

@@

table_name <- "correlations - inner - years"
columns_to_collapse_rows_for <- c("country", "sector", "subsector")
colnames <- c(sectors, "decade", 0L:9L)
align <- make_series_of_repeated_chars(list(l = 4L, r = 10L))
longtable <- TRUE
bold_columns <- 1L:14L
coloured_columns <- 5L:14L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns) %>%
  landscape()
```

@@

table_name <- "regressions - all returns ~ market index"
columns_to_collapse_rows_for <- c("country", "sector", "subsector")
colnames <- c(sectors, periods)
align <- make_series_of_repeated_chars(list(l = 3L, r = 4L))
longtable <- TRUE
bold_columns <- 1L:7L
coloured_columns <- 4L:7L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```

@@

table_name <- "regressions - all returns ~ factors"
columns_to_collapse_rows_for <- c("country", "sector", "subsector", "factor")
colnames <- c(sectors, "factor", "regime", periods)
align <- make_series_of_repeated_chars(list(l = 5L, r = 4L))
longtable <- TRUE
bold_columns <- 1L:9L
coloured_columns <- 6L:9L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns) 
```

\restoregeometry

@@

table_name <- "assets taxonomy"
columns_to_collapse_rows_for <- c("country", "sector", "subsector")
colnames <- c("asset", "country", "sector", "sub-sector")
align <- make_series_of_repeated_chars(list(l = 4L))
longtable <- TRUE
bold_columns <- 1L
coloured_columns <- 2L:4L

make_table(table_name, columns_to_collapse_rows_for, colnames, align, longtable, bold_columns, coloured_columns)
```
